---
title: "Orangutan Spatial Pattern"
author: "Liwen Zhang, Xuechun Liu, Kailun Huang*"
date: "4/11/2018"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
library(sp)
library(GWmodel)
library(maptools)
library(spdep)
library(spgwr)
library(ggplot2)
library(RColorBrewer)
```

### Data Input
```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
OHD1<- rgdal::readOGR("Points_to_use5new.shp")
```


### Introduction
We get this project from our client, Andrea Blackburn, an anthropology student at Boston
University. This project studies how spatial parameters (altitude, slope, distance to river, and
normalized difference vegetation index) predict an orangutan spatial point pattern. The orangutan
point data was collected with handheld GPS units, and the spatial parameter data was extracted
from landsat and DEM imagery using geographic information systems (GIS).
In this project, we use the Kernel Density Value for each point as the dependent variable, and the
geological information of the data points as independent variables, including Normalized
Difference Vegetation Index (NDVI), the degree of the slope (0-90 degrees), elevation measured
in meters, Euclidean distance from point to the river (meters).


### Methods

1. Multiple regression model

First, we run a multiple regression model lm(RASTERVALU ~ NDVI_v2 + slope1 + altitude + dist_riv_b) and look at its residual plot to check model assumptions.

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
### Linear model
model1 <- lm(log(RASTERVALU) ~ NDVI_v2 + slope1 + altitude + dist_riv_b, data = OHD1)
#summary(model1)
#plot(model1, which=3)

plot(model1$fitted.values, residuals(model1), xlab = "Fitted Values", ylab = "Residuals", main = "Residual Plot")
abline(a=-0.41, b=0, col = "red")
abline(a=0.07, b=0, col = "red")
abline(a=0.47, b=0, col = "red")
text(7, -0.41, "Q1", col = "red") 
text(7, 0.07, "Q2", col = "red") 
text(7, 0.47, "Q3", col = "red") 
```

The residual plot reveals a funnel shape, which indicates that there is a violation of heteroscedasticity assumptions. We tried to transform the response Y using a concave function (ex. logY, sqrt(Y)), but the shape wasn't improving much. Another option would be to fit a new model by weighted least squares, which geographically weighted regression(GWR) may be a good choice.

Another way to check is to plot the residuals by its quantile to see if there is any obvious spatial patterning. 

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
resids <- residuals(model1)
colours <- c("dark blue", "blue", "red", "dark red") 
#here it is assumed that your eastings and northings coordinates are stored in columns called x and y in your dataframe
map.resids <- SpatialPointsDataFrame(data=data.frame(resids), coords=cbind(OHD1$Latitude,OHD1$Longitude)) 
#for speed we are just going to use the quick sp plot function, but you could alternatively store your residuals back in your LondonWards dataframe and plot using geom_point in ggplot2
spplot(map.resids, cuts=quantile(resids), col.regions=colours, cex=1, 
       key.space=list(x=1,y=0.9,corner=c(0,1)), 
       xlab = "Longitude", ylab = "Latitude", main = "Residual Spatial Plot") 
```

From this plot it is apparent that there is some spatial patterning of the residuals (i.e. the red, dark red and blue points are not randomly distributed, but there appear to be small clusters of red, dark red and blue points in certain parts of the map). As there appears to be some spatial patterning in these residuals, we will now run a geographically weighted regression (GWR) model to see how the coefficients of the model might vary across space.

2. GWR Model

First we will calibrate the bandwidth of the kernel that will be used to capture the points for each regression and then run the GWR model.

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
#calculate kernel bandwidth
GWRbandwidth <- gwr.sel(RASTERVALU ~ NDVI_v2+slope1+altitude+dist_riv_b, data=OHD1, coords=cbind(OHD1$Latitude,OHD1$Longitude),adapt=T) 
#run the gwr model
gwr.model <- gwr(RASTERVALU ~ NDVI_v2+slope1+altitude+dist_riv_b, data=OHD1, coords=cbind(OHD1$Latitude,OHD1$Longitude), adapt=GWRbandwidth, hatmatrix=TRUE, se.fit=TRUE) 
#print the results of the model
gwr.model
```

The output from the GWR model reveals how the coefficients vary across the Study region. You will see how the global coefficients are exactly the same as the coefficients in the earlier lm model. In this particular model, if we take altitude, we can see that the coefficients range from a minimum value of -19.43 (1 unit change in altitude resulting in a decrease in kernel density value of 19.43) to +19.54 (1 unit change in altitude resulting in an increase in kernel density value of +19.54). For half of the points in the dataset, as altitude rises by 1 meter, kernel density scores will increase between -1.86 and 1.56 points (the inter-quartile range between the 1st Quartile and the 3rd Quartile).

Coefficient ranges can also be seen for the other variables and they suggest some interesting spatial patterning. To explore this we can plot the GWR coefficients for different variables.

```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
# Plot the various GWR coefficients
results <- as.data.frame(gwr.model$SDF)
head(results)
```

In regards of how to interpret the results, let's take the first point as an example. The model we got for the first point after fitting GWR model is $$RASTERVALU = 880.98 + 858.16*NDVI_v2 - 4.78*slope1 + 1.17*altitude - 4.27*dist_riv_b$$. It means that 1 unit change in Normalized Difference Vegetation Index (NDVI) results in an increase in kernel density value of 858.16, when holding other three variables the same and for 1 unit change in slope results in a decrease in kernel density value of 4.78 Also, for 1 unit change in altitude results in an increase in kernel density value of 1.17, while 1 unite change in Euclidean Distance from point to the river in meters results in a decrase of 4.27.

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
#attach coefficients to original dataframe
OHD1$coefNDVI_v2 <- results$NDVI_v2
OHD1$coefslope1 <- results$slope1
OHD1$coefaltitude <- results$altitude
OHD1$coefdist_riv_b <- results$dist_riv_b
OHD2 <- as.data.frame(OHD1)

###NDVI_v2
gwr.point1 <- ggplot(OHD2, aes(x=Longitude,y=Latitude)) +
  geom_point(aes(colour=OHD2$coefNDVI_v2)) +
  scale_colour_gradient2(low = "red", mid = "white", high = "blue", midpoint = 0, 
                         space = "rgb", na.value = "grey50", 
                         guide = "colourbar", guide_legend(title="Coefs"))+
  ggtitle("Coefficient of NDVI")
gwr.point1

###slope1
gwr.point2 <- ggplot(OHD2, aes(x=Longitude,y=Latitude)) +
  geom_point(aes(colour=OHD2$coefslope1)) +
  scale_colour_gradient2(low = "red", mid = "white", high = "blue", midpoint = 0, 
                         space = "rgb", na.value = "grey50", 
                         guide = "colourbar", guide_legend(title="Coefs")) +
  ggtitle("Coefficient of Degree of Slope")
gwr.point2

###altitude
gwr.point3 <- ggplot(OHD2, aes(x=Longitude,y=Latitude)) +
  geom_point(aes(colour=OHD2$coefaltitude)) +
  scale_colour_gradient2(low = "red", mid = "white", high = "blue", midpoint = 0, 
                         space = "rgb", na.value = "grey50", 
                         guide = "colourbar", guide_legend(title="Coefs")) +
  ggtitle("Coefficient of Elevation")
gwr.point3

###dist_riv_b
gwr.point4 <- ggplot(OHD2, aes(x=Longitude,y=Latitude)) +
  geom_point(aes(colour=OHD2$coefdist_riv_b)) +
  scale_colour_gradient2(low = "red", mid = "white", high = "blue", midpoint = 0, 
                         space = "rgb", na.value = "grey50", 
                         guide = "colourbar", guide_legend(title="Coefs")) +
  ggtitle("Coefficient of Euclidean Distance to River")
gwr.point4
```

